---
title: "R4DS ch2"
author: "Zahid Asghar"
date: "5/14/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE)
```

### Remove `eval=FALSE` from above chunk to get results 

## Exercise P#49

```{r}
library(tidyverse)
#1 
library(nycflights13)
#a
glimpse(flights)
flights %>% filter(arr_delay>120) # Arrival delay by more than 2 hours
flights %>% filter(dest=="HOU"|dest=="IAH") # flew to houston
airlines
flights %>% filter(carrier %in% c("AA","DEL","UA"))
flights %>% filter(month==7|month==8|month==9) # in summer
#Or
filter(flights, month %in% 7:9)
flights %>% filter(arr_delay>=120, dep_delay<=0)
filter(flights, dep_delay >= 60, dep_delay - arr_delay > 30)
flights %>% filter(is.na(dep_time))
```

## Exercise  P#51
```{r}
arrange(flights,desc(is.na(dep_delay)))
flights %>% arrange(dep_time) %>% tail() # most delayed
arrange(flights, dep_delay) # departed earliest
head(arrange(flights,desc(distance/air_time))) # fastest
arrange(flights, desc(distance))#longest
arrange(flights, air_time)#shortest
```



## Select()

```{r}
select(flights, year, month, day)
select(flights, year:day) # select all column between year and day (inclusive)
select(flights, -(year:day)) # select all columnsb other than year and day 

```

There are a number of helper functions you can use within select():

`starts_with("abc")`: matches names that begin with “abc”.

`ends_with("xyz")`: matches names that end with “xyz”.

`contains("ijk")`: matches names that contain “ijk”.

`matches("(.)\\1")`: selects variables that match a regular expression. This one matches any variables that contain repeated characters. You’ll learn more about regular expressions in strings.

`num_range("x", 1:3)`: matches x1, x2 and x3.

See ?select for more details.

```{r}
rename(flights,tail_num=tailnum)
select(flights, time_hour,air_time, everything())## Place variables at the start
```


## Exercise P#55


```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)
select(flights, contains("_"))
select(flights, 4,6,7,9)
select(flights, all_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
select(flights, any_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
select(flights, starts_with("dep_"), starts_with("arr_"))
select(flights, contains("_time"), contains("arr_"))
```
### Q2  and 3if variable is selected multiple times

```{r}
select(flights, dep_time, dep_time, arr_time, dep_delay)
vars<-c("year", "month","day","dep_delay","arr_delay")
select(flights,one_of (vars))
select(flights,any_of (vars))
select(flights,all_of (vars))
##4
select(flights, contains("TIME"))
```

## Mutate() to Adding new variable

```{r}
flights_sml<-select(flights, year:day, ends_with("delay"), distance, air_time)
mutate(flights_sml, gain=arr_delay-dep_delay,
       speed=distance/air_time*60)
```
```{r}
## one can refer one has just calculate
mutate(flights_sml,gain=arr_delay-dep_delay,
       hours=air_time/60,
       gain_per_hour=gain/hours)
## To keep only new variables you have created
transmute(flights,gain=arr_delay-dep_delay,
       hours=air_time/60,
       gain_per_hour=gain/hours)

```

## Useful creation function

```{r}
transmute(flights,
          dep_time,
          hour=dep_time %/% 100,
          minute=dep_time %% 100)

```

### Ranking

Look at `row_number()`, `dense_rank()`, `percent_rank()`, `cume_dist()`, and `ntile()`

```{r}
y<-c(1,2,2,NA,3,4)
min_rank(y)
min_rank(desc(y))

row_number(y)
dense_rank(y)
percent_rank(y)
cume_dist(y)

```

## Exercise P#58

```{r}
flights_times <- mutate(flights,
  dep_time_mins = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440,
  sched_dep_time_mins = (sched_dep_time %/% 100 * 60 +
    sched_dep_time %% 100) %% 1440
)
## view only relevant col
select(
  flights_times, dep_time, dep_time_mins, sched_dep_time,
  sched_dep_time_mins
)

```

### Ex2



## `summarise()`

```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```



```{r}
by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE)
)
#> `summarise()` ungrouping output (override with `.groups` argument)
delay <- filter(delay, count > 20, dest != "HNL")

# It looks like delays increase with distance up to ~750 miles 
# and then decrease. Maybe as flights get longer there's more 
# ability to make up delays in the air?
ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)
#> `geom_smooth()` using method = 'loess' and formula 'y ~ x'

```

## Missing values

```{r}
flights %>% group_by(year, month, day) %>% 
  summarise(mean=mean(dep_delay))
## lot of missing values in above command, so
flights %>% group_by(year, month, day) %>% 
  summarize(mean=mean(dep_delay,na.rm=TRUE))
## To handle cancelled flights
not_cancelled<-flights %>% 
  filter(!is.na(dep_delay),!is.na(arr_delay))

not_cancelled %>% group_by(year, month, day) %>% 
  summarise(mean=mean(dep_delay))

```

## Counts

```{r}
# count(n()) or sum(!is.na(x))

delays<-not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(delay=mean(arr_delay))

ggplot(data=delays, mapping = aes(x=delay))+
  geom_freqpoly(binwidth=10)
```

#### For better insight, draw a scatter plot

```{r}
delays<-not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(delay=mean(arr_delay,na.rm=TRUE),
n=n()
)
ggplot(data = delays, mapping = aes(x=n,y=delay))+
  geom_point(alpha=1/10)

```

```{r}
delays %>% 
  filter(n>25) %>% 
    ggplot(mapping = aes(x=n,y=delay))+
  geom_point(alpha=1/10)
```

## Lahman data

```{r}
# Convert to a tibble so it prints nicely
batting<-as_tibble(Lahman::Batting)
batters<-batting %>% 
  group_by(playerID) %>% 
  summarise(ba=sum(H,na.rm = TRUE)/sum(AB,na.rm = TRUE),
            ab=sum(AB,na.rm=TRUE))

```


```{r}
batters %>% 
  ggplot(mapping = aes(x=ab,y=ba))+
  geom_point()+
  geom_smooth(se=FALSE)

```

```{r}
batters %>% arrange(desc(ba))

```

## Useful Summary Functions

```{r}
not_cancelled %>% group_by(year, month, day) %>% 
  summarise(#average delay:
    avg_delay1=mean(arr_delay),
    #average positive delay
    avg_delay2=mean(arr_delay[arr_delay>0]))
    

```


### Measure of spread

```{r}
not_cancelled %>% 
  group_by(distance_sd=sd(distance)) %>% 
  arrange(desc(distance_sd))

```

### Measure of rank min(x), quantile(x,0.25), max(x)

```{r}
#When do the first and last flights leave each day?
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(first=min(dep_time),
            last=max(dep_time))

```

## Counts

```{r}
# which destination has most carriers
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(carriers=n_distinct(carrier)) %>% 
  arrange(desc(carriers))


```



```{r}
not_cancelled %>% 
  count(dest)
not_cancelled %>% count(tailnum,wt=distance)
```
```{r}
# How many flights left before 5am? (these usually indicate delayed
# flights from the previous day)
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(n_early = sum(dep_time < 500))

# What proportion of flights are delayed by more than an hour?
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(hour_prop = mean(arr_delay > 60))
```




## Grouping by Multiple Variables

```{r}
daily<-group_by(flights, year, month, day)
(per_day<-summarise(daily, flights=n()))


```

## Ungrouping

```{r}
daily %>% ungroup() %>%  ## No longer grouped by date
  summarise(flights=n()) ## all flights
```

## Exercises

```{r}

```

## Grouped Mutates (and Filters)


```{r}
flights_sml %>% group_by(year, month, day) %>% 
  filter(rank(desc(arr_delay))<10)

```
## Find all groups bigger than a threshold

```{r}
popular_dest<-flights %>% 
  group_by(dest) %>% 
  filter(n()>365)
popular_dest
```

## Standardize to compute per group metrics

```{r}
popular_dest %>% 
  filter(arr_delay>0) %>% 
  mutate(prop_delay=arr_delay/sum())

```




## Exercise P75

```{r}
flights %>% 
  group_by(carrier) %>% 
  summarise(arr_delay=mean(arr_delay,na.rm=TRUE)) %>% 
  arrange(desc(arr_delay))
filter(airlines,carrier=="F9")
```

