---
title: "ch9 tidyr"
author: "Zahid Asghar"
date: "5/27/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Tidyr

> Happy families are all alike, every unhappy family is unhappy in its own way.
-Leo Tolstoy

> Tidy datasets are all alike, but every messy dataset is messy in its own way.
-Hadley Wickham


```{r}

library(tidyverse)

```

```{r}
table2
t2_cases<-filter(table2, type=="cases") %>% 
  rename(cases=count) %>% 
  arrange(country, cases)
t2_population<-filter(table2, type=="population") %>% 
  rename(population=count) %>% 
  arrange(country, population)

```



```{r}
t2_cases_per_cap<-tibble(
  year=t2_cases$year,
  country=t2_cases$country,
  cases=t2_cases$cases,
  population=t2_population$population
) %>% 
  mutate(cases_per_cap=(cases/population)*1000) %>% 
  select(country,year, cases_per_cap)
t2_cases_per_cap
  

```

```{r}
t2_cases_per_cap <- t2_cases_per_cap %>%
  mutate(type = "cases_per_cap") %>%
  rename(count = cases_per_cap)

```


```{r}

bind_rows(table2, t2_cases_per_cap) %>%
  arrange(country, year, type, count)

```

```{r}
table2 %>% 
  filter(type=="cases") %>% 
  ggplot(aes(year,count))+
  geom_line(aes(group=country),color="grey50")+
  geom_point(aes(color=country))+
  scale_x_continuous(breaks = unique(table2$year))+
  ylab("cases")

```


## Gathering

```{r}
table4a

```

```{r}
table4a %>% 
  gather(`1999`,`2000`,key="year",value = "cases")

```


```{r}
table4b %>% 
  gather(`1999`,`2000`,key="year", value = "population")



tidy4a <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
tidy4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
```



```{r}
stocks<-tibble(
  year=c(2015,2015,2016,2016),
  half=c(1,2,1,2),
  return=c(1.88,0.59,0.92,0.17)
)

stocks %>% 
  pivot_wider(names_from=year,values_from=return) %>% 
  pivot_longer(`2015`:`2016`,names_to = "year", values_to = "return")
```

```{r}
#preg_tidy <- preg %>%
 # pivot_longer(c(male, female), names_to = "sex", values_to = "count")
#preg_tidy
```



## Separating and Pull

```{r}
table3

```

```{r}
table3 %>% 
  separate(rate, into=c("cases","population"))

## Or
table3 %>% 
  separate(rate, into=c("cases","population"),sep = "/")

## Or
table3 %>% 
  separate(rate, into=c("cases","population"),convert = TRUE)


```

```{r}
### One can use last two digits of a year

table3 %>% 
  separate(year,into=c("century","year"),sep=2)
```


## Unite

```{r}

table5 %>% unite(new, century,year)

table5 %>% unite(new, century, year, sep="")
```

## Exercises

```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"))


tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"))

```

```{r}
## Extra drops values without warning while separate drops with warning
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"), extra = "drop")
#`extra` to drop if more values and `fill` if less
```
```{r}
# example with separators
tibble(x = c("X_1", "X_2", "AA_1", "AA_2")) %>%
  separate(x, c("variable", "into"), sep = "_")

```

```{r}
# example with position
tibble(x = c("X1", "X2", "Y1", "Y2")) %>%
  separate(x, c("variable", "into"), sep = c(1))

```

```{r}
# example with separators
tibble(x = c("X_1", "X_2", "AA_1", "AA_2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z])_([0-9])")

```

```{r}
# example with position
tibble(x = c("X1", "X2", "Y1", "Y2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z])([0-9])")

```


Both `separate()` and `extract()` convert a single column to many columns. However, `unite()` converts many columns to one, with a choice of a separator to include between column values.

```{r}
tibble(variable = c("X", "X", "Y", "Y"), id = c(1, 2, 1, 2)) %>%
  unite(x, variable, id, sep = "_")
```


## Missing Values

```{r}
stocks<-tibble(
  year=c(2015,2015,2015,2015,2016,2016,2016),
  qtr=c(1,2,3,4,2,3,4),
  return=c(1.88,0.59,0.35,NA,0.92,0.17,2.66)
)
stocks %>% spread(year,return)
```

```{r}
stocks %>% 
  spread(year, return) %>% 
  gather(year, return, `2015`:`2016`,na.rm=TRUE)
## Another import tool is using `complete()`
stocks %>% 
  complete(year, qtr)
```

```{r}
treatment<-tribble(~person, ~treatment,~response,
                   "Derrick Whitmore",1,7,NA,2,10,NA,
                   3,9,"Katherine Burke",1,4)

treatment %>% 
  fill(person)
```

## Exercises

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
stocks %>% 
  pivot_wider(names_from = year, values_from = return,
              values_fill = 0)
```

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
stocks %>% 
  pivot_wider(names_from = year, values_from = return,
              values_fill = 0)

```
```{r}

stocks %>% 
 complete(year, qtr, fill=list(return=0))

```




## Case Study

```{r}
who
who1<-who %>% 
  gather(new_sp_m014:newrel_f65,key="key",
         value = "cases",
         na.rm = TRUE)
who1
```

#### Get some hint by counting key

```{r}
who1 %>% 
  count(key)

```


```{r}
who2<-who1 %>% 
  mutate(key=stringr::str_replace(key, "newrel","new_rel"))

#who2
who3<-who2 %>% 
  separate(key, c("new","type","sexage"),sep="_")
who3
who3 %>% 
  count(new)
```
```{r}
who4<-who3 %>% 
  select(-new,-iso2,-iso3)

## Next we'll separate sex and age
who5<-who4 %>% 
  separate(sexage,c("sex","age"),sep=1)
who5
## The whole data is now tidy
```

## Complex pipe
Instead of doing steps carried above, one will do as follows

```{r}
who %>% 
  gather(code,value, new_sp_m014:newrel_f65,na.rm = TRUE) %>% 
  mutate(
    code=stringr::str_replace(code,"newrel", "new_rel")
  ) %>% 
  separate(code, c("new","var", "sexage")) %>% 
  select(-new, -iso2, -iso3) %>% 
  separate(sexage, c("sex","age"),sep=1)

```

## Exercises Page#168

```{r}
who1 %>% 
  filter(cases==0) %>% 
  nrow()

```


```{r}
pivot_longer(who, c(new_sp_m014:newrel_f65), names_to = "key", values_to = "cases") %>%
  group_by(country, year) %>%
  mutate(prop_missing = sum(is.na(cases)) / n()) %>%
  filter(prop_missing > 0, prop_missing < 1)

```


```{r}
nrow(who)


who %>% 
  complete(country, year) %>%
  nrow()

```
What happens if mutate(....) step is missed

```{r}
who3a <- who1 %>%
  separate(key, c("new", "type", "sexage"), sep = "_")

filter(who3a, new == "newrel") %>% head()

```

#### Ex4

```{r}
who5 %>%
  group_by(country, year, sex) %>%
  filter(year > 1995) %>%
  summarise(cases = sum(cases)) %>%
  unite(country_sex, country, sex, remove = FALSE) %>%
  ggplot(aes(x = year, y = cases, group = country_sex, colour = sex)) +
  geom_line()
#> `summarise()` regrouping output by 'country', 'year' (override with `.groups` argument)

```








