---
title: "ch12 Factors with forcats"
author: "Zahid Asghar"
date: "6/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Intrdouction


```{r}

library(tidyverse)
library(forcats)
```

## Creating Factors

```{r}
x1<-c("Dec", "Apr", "Jan", "Mar")
x2<-c("Dec","Apr","Jam","Mar")
sort(x1)
```

### Fix both

```{r}
month_levels<-c("Jan","Feb","Mar","Apr","May","Jun", "Jul","Aug",
                "sep","Oct", "Nov","Dec")

y1<-factor(x1,levels = month_levels)
y1
sort(y1)
## Values not in the set will be converted to NA
y2<-factor(x2,levels=month_levels)
y2
```

```{r}
y2<-parse_factor(x2,levels=month_levels)
y2
```


```{r}
f1<-factor(x1,levels=unique(x1))
f1
f2<-x1 %>% factor() %>% fct_inorder()
f2
```

## General Social Survey

```{r}
gss_cat
gss_cat %>% 
  count(race)

```

```{r}
ggplot(gss_cat,aes(race))+
  geom_bar()
```


```{r}
ggplot(gss_cat,aes(race))+
  geom_bar()+
  scale_x_discrete(drop=FALSE)
```
## Exercises 

```{r}

rincome_plot<-ggplot(gss_cat,aes(rincome))+
  geom_bar()
```

```{r}
rincome_plot +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


```{r}
rincome_plot+coord_flip()

```

```{r}
gss_cat %>% 
  filter(!rincome %in% c("Not applicable")) %>% 
  mutate(rincome=fct_recode(rincome,
                             "Less than $1000" = "Lt $1000"
  )) %>% 
   mutate(rincome_na = rincome %in% c("Refused", "Don't know", "No answer")) %>%
  ggplot(aes(x = rincome, fill = rincome_na)) +
  geom_bar() +
  coord_flip() +
  scale_y_continuous("Number of Respondents", labels = scales::comma) +
  scale_x_discrete("Respondent's Income") +
  scale_fill_manual(values = c("FALSE" = "black", "TRUE" = "gray")) +
  theme(legend.position = "None")
```

```{r}
gss_cat %>% 
  count(relig) %>% 
  arrange(desc(n)) %>% 
  head(1)
# most common partyid
gss_cat %>% 
  count(partyid) %>% 
  arrange(desc(n)) %>% 
  head(1)
```


```{r}
gss_cat$denom
gss_cat %>%
  filter(!denom %in% c(
    "No answer", "Other", "Don't know", "Not applicable",
    "No denomination"
  )) %>%
  count(relig)

```
This is also clear in a scatter plot of `relig` vs. `denom` where the points are proportional to the size of the number of answers (since otherwise there would be overplotting).

```{r}
gss_cat %>%
  count(relig, denom) %>%
  ggplot(aes(x = relig, y = denom, size = n)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90))

```


## Modifying Factor Order

```{r}
relig<- gss_cat %>% 
  group_by(relig) %>% 
  summarize(age=mean(age,na.rm=TRUE),
            tvhours=mean(tvhours,na.rm=TRUE),
            n=n())

ggplot(relig,aes(tvhours,relig))+geom_point()

```


```{r}
ggplot(relig,aes(tvhours,fct_reorder(relig,tvhours)))+
  geom_point()

```

For more complicated transformations, it's recommended to move out of `aes()`

```{r}

relig %>% 
  mutate(relig=fct_reorder(relig,tvhours)) %>% 
  ggplot(aes(tvhours,relig))+
  geom_point()

```

```{r}
rincome<-gss_cat %>% 
  group_by(rincome) %>% 
  summarize(age=mean(age,na.rm=TRUE),
            tvhours=mean(tvhours,na.rm=TRUE),
            n=n())
ggplot(rincome,aes(age,fct_reorder(rincome,age)))+
  geom_point()
```


```{r}
ggplot(rincome)+aes(age,fct_relevel(rincome,"Not applicable"))+
  geom_point()
```

`fct_reoder2`
```{r}
by_age<-gss_cat %>% 
  filter(!is.na(age)) %>% 
  group_by(age,marital) %>% 
  count() %>% 
  mutate(prop=n/sum(n))


ggplot(by_age)+aes(age,prop,color=fct_reorder2(marital, age, prop))+
  geom_line()+
  labs(color="marital")
```


#### `fct_infreq`

```{r}
gss_cat %>% 
  mutate(marital=marital %>% fct_infreq() %>% 
           fct_rev()) %>% 
  ggplot(aes(marital))+
  geom_bar()

```


## Exercises

